Turbulence and Particle Acceleration in Giant Radio Haloes: the Origin of Seed Electrons

### 2016-09-12 paper.tex
de-colouring Anders' comments

The bulk of the CRps are injected by relatively low Mach number shocks
and parametrized by $f_\rmn{p,inj}(p) =
C_\rmn{inj}\,p^{-\alpha_\rmn{inj}}$, where $\alpha_\rmn{inj}\approx
2.5$ in our simulations. The CRps approximately trace the thermal gas
with $C_\rmn{inj} \propto \eps_\rmn{th}^{\alpha_\rmn{CR,spat}}$
\citep{pinzke10,2016MNRAS.459...70V}, where the normalisation is fixed
by the \AP{injection rate of} CR energy in the last 650~Myrs. Our simulations show
that \AP{the energy} approximately amounts to 0.03 per cent of the thermal energy inside
the virial radius, i.e.  $\int_0^{R_{200}}\eps_\rmn{CR,inj} \rmn{d}
V\left(\int_0^{R_{200}}\eps_\rmn{th} \rmn{d}V\right)^{-1} =
0.0003$. The spectrum of the initial CRp distribution is determined by
the steady state between injection and cooling,
\begin{equation}
 f_\rmn{p,0} \propto \frac{\int_p^\infty f_\rmn{p,inj}(p') 
\rmn{d}p'}{\displaystyle\left|{{\rmn{d}p}\over{\rmn{d}t}}\right|_{\rm Coul}+\frac{p}{\tau_{\rm had}}}\,,
\end{equation}
where we fix the normalisation by requiring $\int_0^{R_{200}}\eps_\rmn{CR,0}
\rmn{d}V\left(\int_0^{R_{200}}\eps_\rmn{th} \rmn{d}V\right)^{-1} = 0.003$. Note
that the injected energy in the last 650 Myr is smaller by about a factor of 10
in comparison to the cumulative CR energy injected over the entire cosmological
history of the cluster. \AP{However, since the average injection rates of CR energy are similar, 
the CRes injected during the merger are especially important for the radio emission above 1 GHz 
since these CRs have not yet had time to cool.}  Similarly, the initial CRe distribution is given
by the steady state between cooling (Coulomb, inverse Compton, and synchrotron)
and injection of secondary CRes from $f_\rmn{p,0}$.  The diffusion constant
$D_{pp} \propto X_{\rm tu}^{2} \eps_{\rm th}^{\alpha_{\rm tu}-1} \sqrt{T} k_L$ is
calculated for each radial bin. All parameters and assumptions are similar to
what is used for our simulated cluster (see section~\ref{sec:results}). Our
fiducial model assumes $X_{\rm tu} = 0.2$, $\alpha_{\rm tu} =0.8$, $\alpha_{\rm
  CR, spat}=1.0$, and $k_L=2 \upi/\lambda_L$ where $\lambda_L=100$ kpc. We also
assume a fixed acceleration time of $\tau_{\rm cl} = 650 \, {\rm Myr}$.

@@

\C{It is
  interesting to reconsider surface brightness and spectral profiles if indeed
  $\tau_{\rm D} \sim \tau_{\rm cl}$ for the reasons mentioned above (e.g.,
  equation~\ref{eqn:ratios_CR_damp}). We show the results of adopting such an
  ansatz in Figure~\ref{fig:param_comp_tcl_tD}, where, similar to
  Figure~\ref{fig:param_comp}, we vary $X_{\rm tu}$, $\alpha_{\rm tu}$, and
  $\alpha_{\rm CR, spat}$ about fiducial values.\footnote{\C{When \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim$const} is adopted, the radial behaviour of $\tau_{\rm cl}$ is
      also modified. Because $\tau_{\rm D}$ decreases with radius (see
      e.g. Figure~\ref{fig:tauD}), $\tau_{\rm cl}$ has to decrease by the same
      amount, which results in a steeper radio profile. To compensate for this
      effect in Figure~\ref{fig:param_comp_tcl_tD}, we change $\alpha_{\rm CR,
        spat}=1.0\rightarrow0.5$ and \AP{decrease the injection rate of CR energy by a factor five.}}} Instead
  of adopting a fixed $\tau_{\rm cl}$, we set $\tau_{\rm
    cl}/\tau_{\rm D} = 2.5 $. The results are remarkably revealing. In this case, we are indeed less
  sensitive to the properties of the turbulence, and more sensitive to the
  details of the CR seed population.  This gives hope to the possibility that
  one could effectively marginalise over the very uncertain properties of
  turbulence at larger scales in merging clusters (which are unlikely to be more
  precisely constrained observationally in the near future) to learn something
  about the underlying CR population.}

For instance, in the \C{third row of panels} of
Figure~\ref{fig:param_comp_tcl_tD}, we see that once turbulence exceeds a
threshold value $X_{\rm tu} \sim 0.2$, then its overall exact energy density
does not matter -- profiles simply converge to an asymptotic form as $X_{\rm
  tu}$ is increased. This is in contrast to the exponential sensitivity,
flattening in surface brightness, and increasing spectral curvature as $X_{\rm
  tu}$ is increased that we saw previously. A threshold value of $X_{\rm tu}$ is
necessary for turbulent reacceleration to overcome cooling. Once that condition
is satisfied, the condition $\tau_{\rm cl}/\tau_{\rm D} \sim$const implies a fixed
amount of amplification: stronger turbulence implies faster acceleration, but
also dissipates more quickly. This leads to the asymptotic behaviour seen. \C{The value of this constant depends on plasma physics details, but we have shown that the value required to match observations is plausible.}

@@

Finally, in the \C{top panels of Figure~\ref{fig:param_comp_tcl_tD}}, we see
that we {\it are} sensitive to the spatial profile of CRs; it affects both the
radio surface brightness profile (a flatter CR distribution implies flatter
surface brightness profiles) and luminosity (once again, luminosity is larger
for more centrally concentrated CRs, since more secondaries are produced). This
arises because once turbulent reacceleration overcomes cooling, the condition
$\tau_{\rm cl}/\tau_{\rm D} \sim$const provides a fixed amount of amplification in
each radial shell. The radio luminosity in each shell then depends linearly on
initial conditions, i.e., the profile of CR seeds. Interestingly, the
best-fitting profile for Coma results from a flat CR distribution -- i.e., the
outcome of efficient CR streaming. The overall amplitude is set by $\tau_{\rm cl}/\tau_{\rm D}$ (equation
\ref{eqn:ratios_CR_damp}) and the CR content.

@@

\begin{figure*}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{radio profiles}\\
     \includegraphics[width=\columnwidth]{figure7_prof.eps}
   \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{radio spectra}\\
     \includegraphics[width=\columnwidth]{figure7_spec.eps}
   \end{center}
\end{minipage}
\caption{\C{Sensitivity of radio emission in the Coma cluster to critical
    parameters when we vary the ratio of the time scales on which turbulence is
    active ($\tau_{\rm cl}$) to the turbulent re-acceleration time ($\tau_{\rm
      D}$). In comparison to Figure~\ref{fig:param_comp}, we change our fiducial
    model (solid lines) and adopt a slightly more extended spatial distribution
    of CRs ($\alpha_{\rm CR, spat}=0.5$) and lower the \AP{injection rate of CR energy by a factor five}; 
    otherwise, we adopt $X_\rmn{tu}=0.2$, $\alpha_\rmn{tu}=0.8$,
    $\tau_{\rm cl}=2.5\,\tau_{\rm D}$ as fiducial values and vary each parameter
    separately in each row of panels. Radio surface brightness profiles (left)
    are contrasted to radio synchrotron spectra (right; for more details see
    caption to Figure~\ref{fig:param_comp}).  Provided there is a minimum
    (threshold) level of turbulence ($X_\rmn{tu}\gtrsim0.2$), here the radio
    profiles are mainly determined by the CR distribution, while turbulence
    has a \AP{remarkable} small impact. In the bottom two panels we vary the exponential
    momentum growth factor $\tau_{\rm cl}/\tau_{\rm D}$ (see
    equations~\ref{eq:pexp} and \ref{eq:Creacc}), which mainly changes the
    normalisation of the CR distribution and neither spectral shape nor radial
    synchrotron profile. Interestingly, the flat CR profiles adopted here are
    similar to those required to reproduce the radio profiles in our CR
    streaming model (see Figure~\ref{fig:sync_profile}).}}
  \label{fig:param_comp_tcl_tD}
\end{figure*}

@@

If instead we adopt \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim$const}, i.e. the natural
lifetime of turbulence is its decay lifetime, then we can show that the
acceleration time and the lifetime of turbulence are coupled, simply because the
wave-particle interaction rate (which drives particle acceleration) and the
wave-wave interaction rate (which drives dissipation) are linked. If we are
mindful that Kraichnan turbulence is only applicable below the Alfven scale
$l_{\rm A}$ (where turbulent velocities $\varv < \varv_{\rm A}$), then remarkably the
ratio $\tau_{\rm cl}/\tau_{\rm D}$ is independent of properties of the
turbulence (such as its amplitude and outer scale) and only depends on plasma
parameters. Our results suggest that TTD on thermal particles may result in
overly rapid damping of turbulence, and inefficient acceleration (equation
\ref{eqn:ratios_TTD}), but if plasma instabilities scatter the thermal particles
so that the fast modes damp via TTD on the CRs, then the required
acceleration can take place (equation \ref{eqn:ratios_CR_damp}). In agreement
with other assessments, we find that if turbulence is Burgers rather than
Kraichnan, then turbulent reacceleration is ineffective, and some other
explanation for radio haloes (e.g., magnetic reconnection, \citealt{brunetti16}) is
necessary. If we adopt the ansatz \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim$const}, then we
find that above a threshold level of turbulence (necessary to overcome cooling),
radio properties are insensitive to both the amplitude and spatial profile of
turbulence, since only a fixed, asymptotic amount of amplification takes
place. On the other hand, they {\it are} sensitive to the properties of the seed
electron profile. This raises the exciting possibility that one could
marginalise over the highly uncertain properties of cluster turbulence to learn
about the underlying seed CR population. Our results suggest that studying not
just mean trends, but the relatively small scatter in radio-halo scaling
relations, may prove extremely fruitful in shedding light on ICM plasma
processes.

>>>>

The bulk of the CRps are injected by relatively low Mach number shocks and
parametrized by $f_\rmn{p,inj}(p) = C_\rmn{inj}\,p^{-\alpha_\rmn{inj}}$, where
$\alpha_\rmn{inj}\approx 2.5$ in our simulations. The CRps approximately trace
the thermal gas with $C_\rmn{inj} \propto \eps_\rmn{th}^{\alpha_\rmn{CR,spat}}$
\citep{pinzke10,2016MNRAS.459...70V}, where the normalisation is fixed by the
injection rate of CR energy in the last 650~Myrs. Our simulations show that the
CR energy approximately amounts to 0.03 per cent of the thermal energy inside
the virial radius, i.e.  $\int_0^{R_{200}}\eps_\rmn{CR,inj} \rmn{d}
V\left(\int_0^{R_{200}}\eps_\rmn{th} \rmn{d}V\right)^{-1} = 0.0003$. The
spectrum of the initial CRp distribution is determined by the steady state
between injection and cooling,
\begin{equation}
 f_\rmn{p,0} \propto \frac{\int_p^\infty f_\rmn{p,inj}(p') 
\rmn{d}p'}{\displaystyle\left|{{\rmn{d}p}\over{\rmn{d}t}}\right|_{\rm Coul}+\frac{p}{\tau_{\rm had}}}\,,
\end{equation}
where we fix the normalisation by requiring $\int_0^{R_{200}}\eps_\rmn{CR,0}
\rmn{d}V\left(\int_0^{R_{200}}\eps_\rmn{th} \rmn{d}V\right)^{-1} = 0.003$. Note
that the injected energy in the last 650 Myr is smaller by about a factor of 10
in comparison to the cumulative CR energy injected over the entire cosmological
history of the cluster. However, since the individual injection events of CR
energy are similar, the CRes injected during the merger are especially important
for the radio emission above 1 GHz since these CRs have not yet had time to
cool.  Similarly, the initial CRe distribution is given by the steady state
between cooling (Coulomb, inverse Compton, and synchrotron) and injection of
secondary CRes from $f_\rmn{p,0}$.  The diffusion constant $D_{pp} \propto
X_{\rm tu}^{2} \eps_{\rm th}^{\alpha_{\rm tu}-1} \sqrt{T} k_L$ is calculated for
each radial bin. All parameters and assumptions are similar to what is used for
our simulated cluster (see section~\ref{sec:results}). Our fiducial model
assumes $X_{\rm tu} = 0.2$, $\alpha_{\rm tu} =0.8$, $\alpha_{\rm CR, spat}=1.0$,
and $k_L=2 \upi/\lambda_L$ where $\lambda_L=100$ kpc. We also assume a fixed
acceleration time of $\tau_{\rm cl} = 650 \, {\rm Myr}$.

@@

\C{It is interesting to reconsider surface brightness and spectral profiles if
  indeed $\tau_{\rm D} \sim \tau_{\rm cl}$ for the reasons mentioned above
  (e.g., equation~\ref{eqn:ratios_CR_damp}). We show the results of adopting
  such an ansatz in Figure~\ref{fig:param_comp_tcl_tD}, where, similar to
  Figure~\ref{fig:param_comp}, we vary $X_{\rm tu}$, $\alpha_{\rm tu}$, and
  $\alpha_{\rm CR, spat}$ about fiducial values.\footnote{\C{When $\tau_{\rm
        cl}/\tau_{\rm decay} \sim\rmn{const}$ is adopted, the radial behaviour
      of $\tau_{\rm cl}$ is also modified. Because $\tau_{\rm D}$ decreases with
      radius (see e.g. Figure~\ref{fig:tauD}), $\tau_{\rm cl}$ has to decrease
      by the same amount, which results in a steeper radio profile. To
      compensate for this effect in Figure~\ref{fig:param_comp_tcl_tD}, we
      change $\alpha_{\rm CR, spat}=1.0\rightarrow0.5$ and decrease the
      injection rate of CR energy by a factor five.}} Instead of adopting a
  fixed $\tau_{\rm cl}$, we set $\tau_{\rm cl}/\tau_{\rm D} = 2.5 $. The results
  are remarkably revealing. In this case, we are indeed less sensitive to the
  properties of the turbulence, and more sensitive to the details of the CR seed
  population.  This gives hope to the possibility that one could effectively
  marginalise over the very uncertain properties of turbulence at larger scales
  in merging clusters (which are unlikely to be more precisely constrained
  observationally in the near future) to learn something about the underlying CR
  population.}

For instance, in the \C{third row of panels} of
Figure~\ref{fig:param_comp_tcl_tD}, we see that once turbulence exceeds a
threshold value $X_{\rm tu} \sim 0.2$, then its overall exact energy density
does not matter -- profiles simply converge to an asymptotic form as $X_{\rm
  tu}$ is increased. This is in contrast to the exponential sensitivity,
flattening in surface brightness, and increasing spectral curvature as $X_{\rm
  tu}$ is increased that we saw previously. A threshold value of $X_{\rm tu}$ is
necessary for turbulent reacceleration to overcome cooling. Once that condition
is satisfied, the condition $\tau_{\rm cl}/\tau_{\rm D} \sim\rmn{const}$ implies a
fixed amount of amplification: stronger turbulence implies faster acceleration,
but also dissipates more quickly. This leads to the asymptotic behaviour
seen. \C{The value of this constant depends on plasma physics details, but we
  have shown that the value required to match observations is plausible.}

@@

Finally, in the \C{top panels of Figure~\ref{fig:param_comp_tcl_tD}}, we see
that we {\it are} sensitive to the spatial profile of CRs; it affects both the
radio surface brightness profile (a flatter CR distribution implies flatter
surface brightness profiles) and luminosity (once again, luminosity is larger
for more centrally concentrated CRs, since more secondaries are produced). This
arises because once turbulent reacceleration overcomes cooling, the condition
$\tau_{\rm cl}/\tau_{\rm D} \sim\rmn{const}$ provides a fixed amount of amplification in
each radial shell. The radio luminosity in each shell then depends linearly on
initial conditions, i.e., the profile of CR seeds. Interestingly, the
best-fitting profile for Coma results from a flat CR distribution -- i.e., the
outcome of efficient CR streaming. The overall amplitude is set by $\tau_{\rm cl}/\tau_{\rm D}$ (equation
\ref{eqn:ratios_CR_damp}) and the CR content.

@@

\begin{figure*}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{radio profiles}\\
     \includegraphics[width=\columnwidth]{figure7_prof.eps}
   \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{radio spectra}\\
     \includegraphics[width=\columnwidth]{figure7_spec.eps}
   \end{center}
\end{minipage}
\caption{\C{Sensitivity of radio emission in the Coma cluster to critical
    parameters when we vary the ratio of the time scales on which turbulence is
    active ($\tau_{\rm cl}$) to the turbulent re-acceleration time ($\tau_{\rm
      D}$). In comparison to Figure~\ref{fig:param_comp}, we change our fiducial
    model (solid lines) and adopt a slightly more extended spatial distribution
    of CRs ($\alpha_{\rm CR, spat}=0.5$) and lower the injection rate of CR
    energy by a factor five; otherwise, we adopt $X_\rmn{tu}=0.2$,
    $\alpha_\rmn{tu}=0.8$, $\tau_{\rm cl}=2.5\,\tau_{\rm D}$ as fiducial values
    and vary each parameter separately in each row of panels. Radio surface
    brightness profiles (left) are contrasted to radio synchrotron spectra
    (right; for more details see caption to Figure~\ref{fig:param_comp}).
    Provided there is a minimum (threshold) level of turbulence
    ($X_\rmn{tu}\gtrsim0.2$), here the radio profiles are mainly determined by
    the CR distribution, while turbulence has a remarkably small impact. In the
    bottom two panels we vary the exponential momentum growth factor $\tau_{\rm
      cl}/\tau_{\rm D}$ (see equations~\ref{eq:pexp} and \ref{eq:Creacc}), which
    mainly changes the normalisation of the CR distribution and neither spectral
    shape nor radial synchrotron profile. Interestingly, the flat CR profiles
    adopted here are similar to those required to reproduce the radio profiles
    in our CR streaming model (see Figure~\ref{fig:sync_profile}).}}
  \label{fig:param_comp_tcl_tD}
\end{figure*}

@@

If instead we adopt \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim\rmn{const}$},
i.e. the natural lifetime of turbulence is its decay lifetime, then we can show
that the acceleration time and the lifetime of turbulence are coupled, simply
because the wave-particle interaction rate (which drives particle acceleration)
and the wave-wave interaction rate (which drives dissipation) are linked. If we
are mindful that Kraichnan turbulence is only applicable below the Alfven scale
$l_{\rm A}$ (where turbulent velocities $\varv < \varv_{\rm A}$), then
remarkably the ratio $\tau_{\rm cl}/\tau_{\rm D}$ is independent of properties
of the turbulence (such as its amplitude and outer scale) and only depends on
plasma parameters. Our results suggest that TTD on thermal particles may result
in overly rapid damping of turbulence, and inefficient acceleration (equation
\ref{eqn:ratios_TTD}), but if plasma instabilities scatter the thermal particles
so that the fast modes damp via TTD on the CRs, then the required acceleration
can take place (equation \ref{eqn:ratios_CR_damp}). In agreement with other
assessments, we find that if turbulence is Burgers rather than Kraichnan, then
turbulent reacceleration is ineffective, and some other explanation for radio
haloes (e.g., magnetic reconnection, \citealt{brunetti16}) is necessary. If we
adopt the ansatz \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim\rmn{const}$}, then we
find that above a threshold level of turbulence (necessary to overcome cooling),
radio properties are insensitive to both the amplitude and spatial profile of
turbulence, since only a fixed, asymptotic amount of amplification takes
place. On the other hand, they {\it are} sensitive to the properties of the seed
electron profile. This raises the exciting possibility that one could
marginalise over the highly uncertain properties of cluster turbulence to learn
about the underlying seed CR population. Our results suggest that studying not
just mean trends, but the relatively small scatter in radio-halo scaling
relations, may prove extremely fruitful in shedding light on ICM plasma
processes.
